---
title: "Econ 432 HW2 - Tianyu Sun Submission"
output: html_document
---

#Part 1 R Questions 

1. 
```{r}
#Done on paper, see attached picture file in submissions
```


2. 
Suppose X is a normally distributed RV w/ mean 0 and var 1
Compute the following using the R functions pnorm and qnorm 

```{r cars}
#a. Pr(X< -1.96)
p1a = pnorm(-1.96, mean=0, sd=1)
p1a

#b. Pr(X > 1.64)
p1b = pnorm(1.64, mean=0, sd=1)
p1b_answer = 1 - p1b
p1b_answer

#c. Pr(-0.5 < X < 0.5)
p1c = pnorm(-0.5, mean=0, sd=1)
p1c_2 = pnorm(0.5,mean=0,sd=1)
p1c_2
p1c_answer = p1c_2 - p1c
p1c_answer

#d. 1% Quantile (q.01) and 99% Quantile (q.99)
qnorm(.01, mean=0, sd=1)
qnorm(.99,mean=0,sd=1)

#e. 5% Quantile (q.05) and 95% Quantile (q.99)
qnorm(.05,mean=0,sd=1)
qnorm(.99,mean=0,sd=1)

```

3.
Let X denote monthly return on Microsoft Stock and Y is the monthly return for SBUX
Assume X follows N(.05,.1) amd Y follow N(.025,.05)

```{r}
#a. Using a grid of values b/n -.25 and .35, plot the normal curves for X and Y on the same plot 
x = seq(-.25,.35,.01)
f.x <- dnorm(x,mean=.05,sd=.1)
f.y = dnorm(x,mean=.025,sd=.05)

plot(f.y~x, type="l", col="red", ylab="f(x)", main="Density Curves for Microsoft and Starbucks")
lines(f.x~x, type="l", col="blue")

#b. 

#It looks like SBUX is a lot more consistent and has less variance. 
#Microsoft however has a higher average return, but more variance. 
#I would just choose Microsoft as it has higher return and I am also biased as I have Microsoft stocks myself. 

```


4
Let R denote the simple monthly return on Microsoft stock and let W_0 denote initial wealth to be invested over the month. Assume that R follows N(.05,.12) and that W_0 = 100,000

```{r}
#a. Determine the 1% and 5% VaR over the month of the investment 
#1%
library(scales)
W0 <- 100000
alpha <- 0.01
mu.r = 0.05
sigma.r = 0.12

q_r <-  qnorm(alpha, mu.r, sigma.r)
VaR_r <- W0 * q_r
sprintf("The %s VaR is %.3f", percent(alpha), VaR_r)

#5%
W0 <- 100000
alpha <- 0.05
mu.r = 0.05
sigma.r = 0.12

q_r <-  qnorm(alpha, mu.r, sigma.r)
VaR_r <- W0 * q_r
sprintf("The %s VaR is %.3f", percent(alpha), VaR_r)
```
```{r}
#b. Determine the 1% and 5% Expected Shortfall over the month on the investment
#1%
alpha = .01

Integrand <- function(u, alpha_value, mu, sigma) {
  W0*(qnorm(u, mu, sigma))/alpha_value
}
res <- integrate(Integrand, lower = 0, upper = alpha,
                 alpha_value=alpha,
                 mu = mu.r,
                 sigma = sigma.r)
res

sprintf("The %s ES is %#.3f", percent(alpha), res[1])

#5%
alpha = .05

Integrand <- function(u, alpha_value, mu, sigma) {
  W0*(qnorm(u, mu, sigma))/alpha_value
}
res <- integrate(Integrand, lower = 0, upper = alpha,
                 alpha_value=alpha,
                 mu = mu.r,
                 sigma = sigma.r)
res

sprintf("The %s ES is %#.3f", percent(alpha), res[1])
```

5.
```{r}
#a. Determine the 1% and 5% VaR over the month of the investment 
#1%
library(scales)
W0 <- 100000
alpha <- 0.01
mu.r = 0.05
sigma.r = 0.12

q_r <-  qnorm(alpha, mu.r, sigma.r)
VaR_r <- W0 * (exp(q_r)-1)
sprintf("The %s VaR is %.3f", percent(alpha), VaR_r)

#5%
W0 <- 100000
alpha <- 0.05
mu.r = 0.05
sigma.r = 0.12

q_r <-  qnorm(alpha, mu.r, sigma.r)
VaR_r <- W0 * (exp(q_r)-1)
sprintf("The %s VaR is %.3f", percent(alpha), VaR_r)

```
```{r}
#b. 
#Determine the 1% and 5% Expected Shortfall over the month on the investment

alpha = .01

Integrand <- function(u, alpha_value, mu, sigma) {
  W0*(exp(qnorm(u, mu, sigma))-1)/alpha_value
}
res <- integrate(Integrand, lower = 0, upper = alpha,
                 alpha_value=alpha,
                 mu = mu.r,
                 sigma = sigma.r)
res

sprintf("The %s ES is %#.3f", percent(alpha), res[1])

#5%
alpha = .05

Integrand <- function(u, alpha_value, mu, sigma) {
  W0*(exp(qnorm(u, mu, sigma))-1)/alpha_value
}
res <- integrate(Integrand, lower = 0, upper = alpha,
                 alpha_value=alpha,
                 mu = mu.r,
                 sigma = sigma.r)
res

sprintf("The %s ES is %#.3f", percent(alpha), res[1])
```


```{r}
#c.
#Determine the 1% and 5% VaR over the year on the investment 

#1%
mu = 12*mu.r
abs.sd = sqrt(12*(sigma.r)^2)
alpha = .01
VaR.01 = W0*(exp(qnorm(alpha,mu,abs.sd))-1)
sprintf("The %s VaR is %.3f", percent(alpha), VaR.01)

#5%
mu = 12*mu.r
abs.sd = sqrt(12*(sigma.r)^2)
alpha = .05
VaR.05 = W0*(exp(qnorm(alpha,mu,abs.sd))-1)
sprintf("The %s VaR is %.3f", percent(alpha), VaR.05)
```

```{r}
#d.
#Determine the 1% and 5% Expected Shortfall over the year 

#1%
alpha = .01

Integrand <- function(u, alpha_value, mu, sigma) {
  W0*(exp(qnorm(u, mu, sigma))-1)/alpha_value
}
res <- integrate(Integrand, lower = 0, upper = alpha,
                 alpha_value=alpha,
                 mu = mu,
                 sigma = abs.sd)
res

sprintf("The %s ES is %#.3f", percent(alpha), res[1])


#5%
alpha = .05

Integrand <- function(u, alpha_value, mu, sigma) {
  W0*(exp(qnorm(u, mu, sigma))-1)/alpha_value
}
res <- integrate(Integrand, lower = 0, upper = alpha,
                 alpha_value=alpha,
                 mu = mu,
                 sigma = abs.sd)
res

sprintf("The %s ES is %#.3f", percent(alpha), res[1])
```



6. 

```{r}
#a. On the same graph, plot the probability curves of chi-squared distributed 
#random variables with 1, 2, 4 and 10 degrees of freedom.
#Use different colors and line styles for each curve.

x <- seq(-5,20,by=1)


x.c1 <- dchisq(x,df=1)
x.c2 <- dchisq(x,df=2)
x.c4 <- dchisq(x,df=4)
x.c10 <- dchisq(x,df=10)


plot(x.c1~x, type="l", col="blue", 
     lwd=2, ylim=c(0,0.6), xlab = "x", ylab = "f(x)", 
     main = "Density Curves of Multiple Distributions")

lines(x.c2~x, type="l", col="red", 
      lwd=2)

lines(x.c4~x, type = "l", col = "green", lwd = 2)

lines(x.c10~x, type="l", col = "black", lwd = 2)

legend(x="topleft", legend=c("chi(1)","chi(2)","chi(4)","chi(10)"),
       col=c("blue","red", "green","black"), lwd=2)

#b. 
#On the same graph, plot the prob curves of Student t distributed RV w/ 1,2,4,10 df
#Use different colors and line styles for each curve
x.t1 <- dt(x,df=1)
x.t2 <- dt(x,df=2)
x.t4 <- dt(x,df=4)
x.t10 = dt(x,df=10)


lines(x.t1~x, type="l", lty = 2, col="purple", 
      lwd=2)

lines(x.t2~x, type = "l", lty = 2, col = "orange", lwd = 2)

lines(x.t4~x, type="l", lty = 2, col = "brown", lwd = 2)

lines(x.t4~x, type="l", lty = 2, col = "yellow", lwd = 2)

legend(x = "topright", legend = c("t(1)","t(2)","t(4)","t(10)"), col = c("purple","orange","brown","yellow"), lwd=2)
```




```{r}
#c. 
#Without doing any calculation, which one do you expect is greater in ab value between 
#5% VaR of standard normal vs 5% VaR of t distribution w/ df=2


#I would say that the VaR of the t distribution w/ df=2 would be higher 
```



7.
```{r}
#Done on Paper, see attached picture file in submissions

```



Part 2 R Questions 

1.

```{r}
#Suppose that X is a normal random variable with mean 0:1
#and variance 1. Using Monte Carlo simulation to Önd the probability that X>=0.5
set.seed(123) 
runs <- 5000
mu.x <- 0.1
sigma.x <- 1

X <- rnorm(runs, mu.x, sigma.x) 
sum(X>=0.5)/runs
```



2. 
```{r}
#Flip a coin, you and a friend guess the outcome H or T
#If you both guess the same, lose 3 if H, lose 1 if T
#If guess is different from your friend, you win 2
#You both observe the prob of observing H is 0.5
#What is the expectation of the money you will win from this game?

set.seed(321)
X2 <- matrix(rbinom(runs*2,size = 1, prob = 0.5), ncol = 2)
head(X2)

Y <- ifelse(X2[,1]==1 & X2[,2]==1, -3,
            ifelse(X2[,1]==0 & X2[,2]==0 , -1, 2))
head(cbind(X,Y)) 

sum(Y)/runs
```



3. 
```{r}
#Consider the game described in #2, Now you play the game w/ Prob(X=1)=0.36
#Suppose your friend still has Prob(X=1)=0.5
#What is the expectation of the money you will win from this game?

X3_Person1 =  matrix(rbinom(runs*2,size = 1, prob = 0.36), ncol = 2)
head(X3_Person1)

X3_Person2 = matrix(rbinom(runs*2,size = 1, prob = 0.5), ncol = 2)
head(X3_Person2)

Y <- ifelse(X3_Person1[,1]==1 & X3_Person2[,2]==1, -3,
            ifelse(X3_Person1[,1]==0 & X3_Person2[,2]==0 , -1, 2))
head(cbind(X,Y)) 

sum(Y)/runs
```



4. 
```{r}
#Consider the game in #2. Mathematically, we can show that strategy in #3 
#Guarantees that you will win money from your friend regardless the guess of your friend 
#If you play the games many times with your friend. That is, we can show that 
#the expectation of the money you get from the  game is positive regardless of
#your friends guess. Check this by simulation


sum(Y)/runs
  #It has a positive expected value 

```

 





















